A simulation model of DNA methylation data

### 2014-10-19 Rmarkdown/avy.Rmd
Section-level tidying

# Chapter overview

The agouti viable yellow ($A^{vy}$) mouse model is the archetypical _epiallele_, a genetically identical region where epigenetic variability result in phenotypic variation. The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins the $A^{vy}$ model, recently performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ locus, with the aim of identifying additional epialleles.

@@

# Introduction + literature review

@@

## A general strategy for identifying epialleles of known genes

@@

# Description of experiment

@@

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Sample code & Mouse number & Agouti genotype           \\ \midrule
sample1     & mouse \#56   & $A^{vy}/a$ (yellow)       \\
sample2     & mouse \#57   & $a/a$                     \\
sample3     & mouse \#58   & $a/a$                     \\
sample4     & mouse \#65   & $A^{vy}/a$ (pseudoagouti) \\
sample5     & mouse \#66   &                           \\ \bottomrule
\end{tabular}
\caption{Description of mice in Whitelaw experiment.}
\label{tab:whitelaw_data}
\end{table}

@@

## Data

@@

# Methodology

@@

This is the first example I had seen of a bisulfite-sequencing experiment with one-group. Surprisingly, this rules out all(?) published methods for identifying differential methylation, since these are all designed for identifying differential methylation in experiments with two or more groups.

There is also the question of what sized "features" we are interested in testing for differential methylation. Harald and Emma told me that they were interested in both differentially methylated CpGs and differentially methylated regions, where the regions were not defined _a priori_. I took the commonly used approach of testing individual CpGs for differential methylation and then building runs of DMCs to define differentially methylated regions.

My algorithm for calling differentially methylated regions is adapted from that of \cite{Hansen:2011gu}. I call my implementation `runDMC` (runs of differentially methylated cytosines) and it proceeds as follows:

@@

## Step 1: Identifying candidate differentially methylated cytosines

@@

\begin{equation*}

H_{0}:& \beta_{i, j} = \beta_{i, 0}, j = 1, \ldots, 5 \\
  vs.& \\
H_{1}:& \exists j = 1, \ldots, 5 s.t. \beta_{i, j} \neq \beta_{i, 0}

\end{equation*}

@@

The question of

## Identifying differentially methylated regions

@@

`runDMC` is written in `C++` and interfaces with `R` via the `Rcpp` package (__CITE__). The `C++` version is some $25$ times faster than the equivalent code written in `R` (__TODO: Include benchmark data__). The user passes it an object containing the set of CpGs, each annotated by whether it is a _cDMC_, and specifies __UP TO HERE__

# Results

# Discussion

# Conclusion

>>>>

## Chapter overview

The agouti viable yellow ($A^{vy}$) allele is the archetypical _epiallele_, a genetically identical region where epigenetic variability result in phenotypic variation. The research group of Professor Emma Whitelaw, one of the researchers to discover the epigenetic variant that underpins the $A^{vy}$ mouse model,  performed whole-genome bisulfite-sequencing of five mice, including two with the $A^{vy}$ allele, with the aim of identifying additional epialleles.

@@

## Introduction + literature review

@@

### A general strategy for identifying epialleles of known genes

@@

## Description of experiment

@@

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Sample code & Mouse number & Agouti genotype           \\ \midrule
sample1     & mouse \#56   & $A^{vy}/a$ (yellow)       \\
sample2     & mouse \#57   & $a/a$                     \\
sample3     & mouse \#58   & $a/a$                     \\
sample4     & mouse \#65   & $A^{vy}/a$ (pseudoagouti) \\
sample5     & mouse \#66   &                           \\ \bottomrule
\end{tabular}
\caption{Description of mice in Whitelaw experiment.}
\label{tab:whitelaw_data}
\end{table}

@@

### Data

@@

## Methodology

@@

This is the first example I had seen of a bisulfite-sequencing experiment with one-group. Surprisingly, this rules out all(?) published methods for identifying differential methylation, since these are all designed for identifying differential methylation in experiments between two or more groups.

There is also the question of what sized "features" we are interested in testing for differential methylation. Harald and Emma told me that they were interested in both differentially methylated CpGs and differentially methylated regions, where the regions were not defined _a priori_. For defining DMRs I took the commonly used approach of testing individual CpGs for differential methylation and then defining DMRs as runs of DMCs.

My algorithm for defining differentially methylated regions is adapted from that of \cite{Hansen:2011gu}. I call my implementation `runDMC` (runs of differentially methylated cytosines) and it proceeds as follows:

@@

### Step 1: Identifying candidate differentially methylated cytosines

@@

$$
\begin{aligned}
H_{0, i}: & \forall j \; B_{i, j} = B_{i, 0} \\
& vs.  \\
H_{1, i}: & \exists j: \; B{i, j} \neq B_{i, 0}
\end{aligned}
$$

@@

__TODO: Some summaries of the cDMCs, e.g., top-ranked CpGs__

__TODO: A discussion of FDR correction for single-loci tests__

### Identifying differentially methylated regions

@@

`runDMC` is written in `C++` and interfaces with `R` via the `Rcpp` package (__CITE__). The `C++` version is some $25$ times faster than the equivalent code written in `R` (__TODO: Include benchmark data__). The user passes `runDMC` an object containing the set of CpGs, each annotated by whether it is a _cDMC_, and specifies the maximal $IPD$ of CpGs and number of _non-cDMCs_ before a run is "broken". The output is an object with the start and end co-ordinates of each run, as well as some basic summary statistics such as the number of _cDMCs_ and _non-cDMCs_. These cDMRs are then annotated and filtered in the final stage.

__TODO: Some summaries of the `runDMC` output, e.g., number of runs, lengths,

### Summarising and filtering candidate differentially methylated regions

The vast majority of "runs" returned by `runDMC` are a single _cDMC_. Obviously we want to filter these out as they are single loci not runs. For example, we may require a run to contain at least 3 _cDMCs_.

We may want to summarise each _cDMR_ by a statistic that measures how different the entire region is between the sample, not just the individual CpGs. However, care should be taken in the interpretation of such a statistic. We know _a priori_ that the CpGs in an cDMR are 'significantly' differentially methylated and so we expect the region to also be 'significantly' differentially methylated. Nevertheless, a summary statistic may be useful in ranking candidate DMRs.

One potential summary statistic is the Mantel-Haenszel (_MH_) statistic of the CpGs in the _cDMR_. The Mantel-Haenszel statistic can be used to test that two nominal variables are conditional independent in each stratum. In other words,

> The Mantel-Haenszel test tests the null hypothesis that two nominal variables [M and U] are conditionally independent in each stratum [at each CpG], assuming that there is no three-way interaction [assuming that the effect is consistent across all CpGs in the region] (From the help page for the `R` implementation of the Mantel-Haenszel test (see ``?mantelhaen.test`) [with my comments in square brackets])

Testing the conditional independence of $M$ and $U$ _at a single CpG_ is equivalent to testing that $\beta_{j} = \beta_{0}$ for all $j = 1, \ldots, 5$ mice at that CpG. Therefore, the Mantel-Haenszel test is equivalent to testing  the null hypothesis that all mice have the same methylation level at each CpG ($\beta_{0, i}$) in the region, although this may vary from CpG to CpG, i.e. $\beta_{0, i}$ need not equal $\beta_{0, i + 1}$).

In our case, with 2 categories ($M$ and $U$) and 5 mice, the Mantel-Haenszel statistic has an approximate $\chi^{2}((5 - 1) \times (2 - 1) \times(2 - 1)) = \chi^{2}(4)$ distribution under the null hypothesis of no conditional independence of $M$ and $U$ across the sites[^df]. Furthermore, the Mantel-Haenszel statistic is the optimal statistic for the alternative hypothesis stating a consistent (same direction) difference across all CpGs in the region \cite{Somes:1986uk}. We may with to rank regions based on the value of the Mantel-Haenszel statistic.

[^df]: If only a $J < 5$ of the mice have non-zero coverage across all CpGs in the region, then the degrees of freedom are $J - 1$.

There are some important caveats mentioned above that warrant further discussion. One of these, the assumption of no three-way interaction, which is required for the validity of the null distribution of the Mantel-Haenzel statistic, is related to the idea of "consistency" for a DMR.

By "consistency" I mean that the difference in methylation across the region occurs within the same (subset) of mice. This is best illustrated by describing an "inconsistent" DMR. Suppose we have a region containing five CpGs, each of which is differentially methylated. However, suppose that at the first CpG it is `sample1` that is "different", at the second CpG it is `sample2` that is "different", etc. Then this region is not differentially methylated in any single subset of mice, which is the standard interpretation of a DMR. It must be emphasised that the `runDMC` algorithm provides no guarantees that a _cDMR_ is consistent since it simply builds runs of _cDMCs_.

Returning to the Mantel-Haenszel statistic, the assumption of no three-way interaction means no interaction between mouse, methylation level and CpG across the region. Another way to think of this is that no three-way interaction means that there is no CpG-to-CpG heterogeneity across the region in the interaction between mouse and methylation level, i.e., that any difference in methylation between the mice is _consistent_ across the CpGs. __TODO: Discuss interpretation with Terry__.

We could of course test whether there is evidence of a three-way interaction. Specifically, we could fit a log-linear model to the tables in each region and compare the "full" model (mouse $\times$ CpG $\times$ methylation level) against the reduced "all main effects and two-way interactions" model (mouse $\times$ CpG + mouse $\times$ methylation level + CpG $\times$ methylation level) via a likelihood ratio test. Again, we do not use the P-values to strictly rule regions as "significant" or "non-significant", but rather to highlight regions  that may be inconsistent.

__TODO: Is this what we want?__

```{r, eval = FALSE, echo = FALSE}
# Testing whether no 3-way interaction is the same thing as consistency

# Some data where each CpG is sig but region is not consistent
x <- matrix(c(3, 40, 30, 40, 30, 40, 30, 40, 30, 40,
              30, 40, 3, 40, 30, 40, 30, 40, 30, 40,
              30, 40, 30, 40, 3, 40, 30, 40, 30, 40,
              30, 40, 30, 40, 30, 40, 3, 40, 30, 40,
              30, 40, 30, 40, 30, 40, 30, 40, 3, 40),
              ncol = 10)

tx <- array(x, dim = c(2, 5, 5))

# Every table is significant
for (i in 1:5) {
  print(chisq.test(tx[, , i]))
}

# MH test is non-significant
mantelhaen.test(tx)

# But that's because no three-way interaction assumption is violated?
llx <- loglin(table = tx, margin = list(c(1, 2), c(1, 3), c(2, 3)))
1 - pchisq(llx$lrt, llx$df)

# Try with some permuted data: Every CpG is DMC and region is consistent
y <- rbind(rep(c(3, 40), 5), matrix(rep(c(30, 40), 20), ncol = 10))

ty <- array(y, dim = c(2, 5, 5))

# Every table is significant
for (i in 1:5) {
  print(chisq.test(ty[, , i]))
}

# MH test is highly significant
mantelhaen.test(ty)

# And there's no 3-way interaction
lly <- loglin(table = ty, margin = list(c(1, 2), c(1, 3), c(2, 3)))
1 - pchisq(lly$lrt, lly$df)
```

## Results

## Discussion

* Analysis doesn't incorporate co-methylation
* Consistency
* Experiment is a bit of a fishing expedition?

## Conclusions

### 2014-10-19 Rmarkdown/methsim.Rmd
Section-level tidying

# Chapter overview

# Introduction

@@

# Background + literature review

@@

## Methods for simulating bisulfite-sequencing reads

@@

## Methods for simulating aggregate methylation levels

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section. Generally, the simulation method is not a major feature of the paper but is there to support claims about the performance of the proposed method. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, of the methods reviewed in this section, only that of \citet{Lacey:2013iy} has a software implementation available.

@@

### Simulating sequence depth

@@

### Simulating $B_{i, j}$

@@

### Simulating co-methylation

@@

As demonstrated in Chapter \ref{chap:co-methylation}, the physical process of co-methylation is at the level of individual DNA fragments, not at the level of aggregate methylation levels. The observed correlations of $\beta$-values are driven by this within-fragment co-methylation. However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments. Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation but it may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependence. This idea takes its inspiration from \citet{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.

@@

### Simulating positions of methylation loci

@@

### Other model-based simulations

@@

## Methods based entirely on sampling real data

@@

## Simulating the truth

@@

### Segmenting the methylome

@@

### Simulating a single haploid methylome

@@

The above-described model, being spatially inhomogeneous, is not particularly amenable to analytical calculations. However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome. For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}$, to estimate or otherwise assign.

@@

### Simulating a single sample

Given a set of transition matrices, $\Big { P_{i^{(h, j)} \Big }$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$. Naively, we might simulate a single sample by simply generating a realisation for each $h$ of $Z_{h, j}$. However, although there is cell-to-cell heterogeneity in the methylation patterns, there is still a degree of dependence for the $i^{th}$ methylation state across the $h$ haploid genomes in the $j^{th}$ sample. After all, the cells are from the same individual and are broadly of the same "type".

What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big { P_{i^{(h, j)} \Big }$, and for those for haploid methylome $h'$, $\Big { P_{i^{(h', j)} \Big }$. This is asking quite a lot of our simple simulation model. Instead, `methsim` uses a heuristic procedure to achieve a similar result.

To begin, we assume that all $H$ haploid genomes in the sample have the same co-methylation structure. That is equivalent to defining $\Big { P_{i^{(h, j)} \Big } := $\Big { P_{i^{(j)} \Big }$ for all $h$. __TODO: Comment on this assumption__

Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big { P_{i^{(j)} \Big }$. However, this gives no control over the dependence between the methylation states at the $i^{th}$ locus over the $H$ haploid genomes in the $j^{th}$ sample. In fact, such a model will have the "correct" co-methylation structure but "incorrect" $\beta$-values. __TODO: Check this and demonstrate using `methsim`__

Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big { P_{i^{(j)} \Big }$. It then weights these realisations according to the observed heterogeneity for each region. __TODO: Describe the inuition of this weighting scheme as being that $w_{r}$ of the cells have identical methylation states for that region.

### Simulating multiple samples from the same group

### Simulating differential methylation

## Parameter estimation

# Implementation

# Case study

# Discussion of findings

# Conclusions

>>>>

## Chapter overview

## Introduction

@@

## Background + literature review

@@

### Methods for simulating bisulfite-sequencing reads

@@

### Methods for simulating aggregate methylation levels

Most papers that propose a new method for downstream analysis of bisulfite-sequencing data include a simulation section. Generally, the simulation method is not a major feature of the paper but is there to support claims about the performance of the proposed method. Consequently, the simulation model is often only briefly described and a software implementation is rarely made available. In fact, of the methods reviewed in this section, only that of \cite{Lacey:2013iy} has a software implementation available.

@@

#### Simulating sequence depth

@@

#### Simulating $B_{i, j}$

@@

#### Simulating co-methylation

@@

As demonstrated in Chapter \ref{chap:co-methylation}, the physical process of co-methylation is at the level of individual DNA fragments, not at the level of aggregate methylation levels. The observed correlations of $\beta$-values are driven by this within-fragment co-methylation. However, a simulation model of DNA methylation data need not induce correlations at the level of individual DNA fragments. Indeed, if the model does not simulate the individual fragments then it cannot simulate within-fragment co-methylation but it may be able to capture correlations of methylation levels by inducing dependence in the $B_{i, j}$. This could be induced under the Beta-Binomial model by forcing the $\mu_{i, j}$ to be spatially dependence. This idea takes its inspiration from \cite{Jaffe:2012gx} who simulate spatially correlated DNA methylation microarray data by imposing an autocorrelation structure [AR(1)] on the simulated $\beta$-values.

@@

#### Simulating positions of methylation loci

@@

#### Other model-based simulations

@@

### Methods based entirely on sampling real data

@@

### Simulating the truth

@@

#### Segmenting the methylome

@@

#### Simulating a single haploid methylome

@@

The above-described model, being spatially inhomogeneous, is not particularly amenable to analytical calculations. However, it is very simple to simulate realisations from this model, requiring only a single loop over the set of loci on each chromosome. For a chromosome containing $n$ methylation loci there are $n - 1$ transition matrices, $P_{i^{(h, j)}}$, to estimate or otherwise assign.

@@

#### Simulating a single sample

Given a set of transition matrices, $\Big \{ P_{i^{(h, j)}} \Big \}$, we can readily simulate a realisation of the methylation states along a haploid genome, $Z_{h, j}$. Naively, we might simulate a single sample by simply generating a realisation for each $h$ of $Z_{h, j}$. However, although there is cell-to-cell heterogeneity in the methylation patterns, there is still a degree of dependence for the $i^{th}$ methylation state across the $h$ haploid genomes in the $j^{th}$ sample. After all, the cells are from the same individual and are broadly of the same "type".

What this means is that we want some degree of dependence between the set of transition matrices for haploid methylome $h$, $\Big \{ P_{i^{(h, j)}} \Big \}$, and for those for haploid methylome $h'$, $\Big \{ P_{i^{(h', j)}} \Big \}$. This is asking quite a lot of our simple simulation model. Instead, `methsim` uses a heuristic procedure to achieve a similar result.

To begin, we assume that all $H$ haploid genomes in the sample have the same co-methylation structure. That is equivalent to defining $\Big \{ P_{i^{(h, j)}} \Big \} := \Big \{ P_{i^{(j)}} \Big \}$ for all $h$. __TODO: Comment on this assumption__

Naively, we might simulate a sample by simulating $H$ realisations of the Markov chain with transition matrices $\Big \{ P_{i^{(j)}} \Big \}$. However, this gives no control over the dependence between the methylation states at the $i^{th}$ locus over the $H$ haploid genomes in the $j^{th}$ sample. In fact, such a model will have the "correct" co-methylation structure but "incorrect" $\beta$-values. __TODO: Check this and demonstrate using `methsim`__

Instead, `methsim` simulates $R << H$ realisations of the Markov chain with transition probabilities given by $\Big \{ P_{i^{(j)}} \Big \}$. It then weights these realisations according to the observed heterogeneity for each region. __TODO: Describe the inuition of this weighting scheme as being that $w_{r}$ of the cells have identical methylation states for that region.

#### Simulating multiple samples from the same group

#### Simulating differential methylation

### Parameter estimation

## Implementation

## Case study

## Discussion of findings

## Conclusions
